---
title: "Comparative Cox Analysis-R conference 2023"
author: "Che Muhammad Nur Hidayat Che Nawi"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: rmdformats::material
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, include=FALSE}
# Prepare environment and load data

library(gtsummary)
library(tidyverse)

stroke <- read.csv("stroke.csv")
str(stroke)

str_mod <- read.csv("stroke.csv")
str(str_mod)
```

```{r, include=FALSE}
# Data wrangling

## For descriptive, status can be factor

# change numeric to factor
stroke$age = as.numeric(stroke$age) 
stroke$dur_days = as.numeric(stroke$dur_days)
stroke$dur_month = as.numeric(stroke$dur_month)
stroke$status = as.numeric(stroke$status)
stroke$gcs_reduct = as.numeric(stroke$gcs_reduct)


stroke <- stroke %>% mutate_if(is.integer, ~as.factor(.)) %>% 
  select(age, dur_days, dur_month, status, gcs_reduct,everything())

# Change factor labels of the levels====

stroke$ss <- factor(stroke$ss, labels = c("HRPZ", "HSNZ", "HSP", "HUSM", "PPUKM"))
stroke$sex <- factor(stroke$sex, labels = c("Female", "Male"))
stroke$ethnicity <- factor(stroke$ethnicity, labels = c("Malay", "Chinese", "Indian", "Others"))
stroke$married <- factor(stroke$married, labels = c("Married", "Single", "Others"))
stroke$dm <- factor(stroke$dm, labels = c("No", "Yes"))
stroke$hpt <- factor(stroke$hpt, labels = c("No", "Yes"))
stroke$ckd <- factor(stroke$ckd, labels = c("No", "Yes"))
stroke$af <- factor(stroke$af, labels = c("No", "Yes"))
stroke$hf_ihd <- factor(stroke$hf_ihd, labels = c("No", "Yes"))
stroke$lipid <- factor(stroke$lipid, labels = c("No", "Yes"))
stroke$smoke <- factor(stroke$smoke, labels = c("No", "Yes"))
stroke$who <- factor(stroke$who, labels = c("Ischemic stroke", "Non ischemic stroke"))
stroke$nihss <- factor(stroke$nihss, labels = c("No stroke symptoms (0)", 
                                                        "Minor stroke (1-4)",
                                                        "Moderate stroke (5-15)",
                                                        "Moderate to severe stroke (16-20)",
                                                        "Severe stroke (21-42)"))
stroke$iv_thrombolysis <- factor(stroke$iv_thrombolysis, labels = c("No", "Yes"))
stroke$iv_thrombectomy <- factor(stroke$iv_thrombectomy, labels = c("No", "Yes"))
stroke$mrs <- factor(stroke$mrs, labels = c("mRS score 0", "mRS score 1",
                                                    "mRS score 2", "mRS score 3",
                                                    "mRS score 4", "mRS score 5",
                                                    "mRS score 6"))
stroke$status <- factor(stroke$status, labels = c("Censored", "Death"))
glimpse(stroke)
```


```{r, include=FALSE}
# Data wrangling

## For survival analysis, status cannot be factor

# change numeric to factor
str_mod$age = as.numeric(str_mod$age) 
str_mod$dur_days = as.numeric(str_mod$dur_days)
str_mod$dur_month = as.numeric(str_mod$dur_month)
str_mod$status = as.numeric(str_mod$status)
str_mod$gcs_reduct = as.numeric(str_mod$gcs_reduct)


str_mod <- str_mod %>% mutate_if(is.integer, ~as.factor(.)) %>% 
  select(age, dur_days, dur_month, status, gcs_reduct,everything())

# Change factor labels of the levels====

str_mod$ss <- factor(str_mod$ss, labels = c("HRPZ", "HSNZ", "HSP", "HUSM", "PPUKM"))
str_mod$sex <- factor(str_mod$sex, labels = c("Female", "Male"))
str_mod$ethnicity <- factor(str_mod$ethnicity, labels = c("Malay", "Chinese", "Indian", "Others"))
str_mod$married <- factor(str_mod$married, labels = c("Married", "Single", "Others"))
str_mod$dm <- factor(str_mod$dm, labels = c("No", "Yes"))
str_mod$hpt <- factor(str_mod$hpt, labels = c("No", "Yes"))
str_mod$ckd <- factor(str_mod$ckd, labels = c("No", "Yes"))
str_mod$af <- factor(str_mod$af, labels = c("No", "Yes"))
str_mod$hf_ihd <- factor(str_mod$hf_ihd, labels = c("No", "Yes"))
str_mod$lipid <- factor(str_mod$lipid, labels = c("No", "Yes"))
str_mod$smoke <- factor(str_mod$smoke, labels = c("No", "Yes"))
str_mod$who <- factor(str_mod$who, labels = c("Ischemic stroke", "Non ischemic stroke"))
str_mod$nihss <- factor(str_mod$nihss, labels = c("No stroke symptoms (0)", 
                                                        "Minor stroke (1-4)",
                                                        "Moderate stroke (5-15)",
                                                        "Moderate to severe stroke (16-20)",
                                                        "Severe stroke (21-42)"))
str_mod$iv_thrombolysis <- factor(str_mod$iv_thrombolysis, labels = c("No", "Yes"))
str_mod$iv_thrombectomy <- factor(str_mod$iv_thrombectomy, labels = c("No", "Yes"))
str_mod$mrs <- factor(str_mod$mrs, labels = c("mRS score 0", "mRS score 1",
                                                    "mRS score 2", "mRS score 3",
                                                    "mRS score 4", "mRS score 5",
                                                    "mRS score 6"))

glimpse(str_mod)
```

# Participant characteristics

- We describe data as mean (SD) and frequency (%) when appropriate
- To make comparisons, we employed statistical tests including the Wilcoxon rank sum test, Pearson's Chi-squared test, and Fisher's exact test.

```{r, echo=FALSE}
# Descriptive analysis by gt summary

## By status

library(gtsummary)

# Choose and rearrange variables
stroke2 <- stroke %>% gtsummary::select(age, ss, sex, ethnicity, married, dm, hpt, ckd, af, hf_ihd, lipid, smoke, who, gcs_reduct, nihss, mrs, iv_thrombolysis, iv_thrombectomy, dur_month, status)

desc <- tbl_summary(stroke2,
                    by = status,
                    statistic = list(all_continuous() ~ "{mean} ({sd})",                                                               all_categorical() ~ "{n} ({p}%)"),
                    digits = list(all_continuous() ~ 2,                                                                             all_categorical() ~ 1),
                    label = list(age ~ "Age", 
                                 ss ~ "Study sites",
                                 sex ~ "Gender",
                                 ethnicity ~ "Ethnicity",
                                 married ~ "Marital status",
                                 dm ~ "Diabetes mellitus, Yes",
                                 hpt ~ "Hypertension, Yes",
                                 ckd ~ "Chronic Kidney Disease, Yes",
                                 af ~ "Atrial fibrillation, Yes",
                                 hf_ihd ~ "Heart disease, Yes",
                                 lipid ~ "Hyperlipidemia, Yes",
                                 smoke ~ "Smoking status, Yes",
                                 who ~ "Diagnosis (WHO)",
                                 gcs_reduct ~ "GCS reduction",
                                 nihss ~ "NIH stroke scale (NIHSS)",
                                 mrs ~ "Modified rankin scale (mRS)",
                                 iv_thrombolysis ~ "Intervention (Thrombolysis), Yes",
                                 iv_thrombectomy ~ "Intervention (Thrombectomy), Yes",
                                 dur_month ~ "Time interval in month"))
desc %>%
  add_p %>% 
  add_overall() %>% 
  bold_labels() %>% 
  italicize_levels() 
```

Table 1: Basic characteristics of the included stroke patients

- The mean (SD) age at the time of diagnosis was 63.15 (13.09) years. 
- The distribution of patients across study centers was fairly even. 
- The majority of patients were male, comprising 552 individuals (58.1%), and of Malay ethnicity, totaling 776 patients (81.7%). 
- Ischemic stroke was the most prevalent type, accounting for 870 cases (91.6%). 
- Less than 10% of the patients received intravascular intervention, with 80 patients (8.4%) undergoing thrombolysis and 61 patients (6.4%) receiving thrombectomy.

# 3-months, 12-months and 36-months stroke survival rate among stroke patients 

Now we will describe the survival based on short term (3month), intermediate term (1year) and long term (3years) survival post stroke. 

```{r, include=FALSE}
## Load Library
library(survival)
library(tidyverse)
library(survminer)
library(scales)
```


```{r, echo=FALSE, warning=FALSE, message=FALSE}
fit1 <- survfit(Surv(dur_month, status)~ 1, str_mod)
fit2 <- survfit(Surv(dur_month, status)~ sex, str_mod)
fit3 <- survfit(Surv(dur_month, status)~ ethnicity, str_mod)
fit4 <- survfit(Surv(dur_month, status)~ who, str_mod)


life_table <- list(fit1, fit2, fit3, fit4) %>% 
  tbl_survfit(times= c(1, 12, 36)) %>% 
  modify_header(update = list(
    stat_1 ~ "**3-month survival (95% CI)**",
    stat_2 ~ "**1-year survival (95% CI)**",
    stat_3 ~ "**3-years survival (95% CI)**"
  )) %>% add_n() %>% add_p() %>% 
  bold_labels() %>% 
  italicize_levels()

life_table
```

Table 3: Overal, gender, ethnicity, and stroke diagnosisl survival rate for stroke patients at 3-month, 12-month, and 36-month

- The proportion of stroke patients alive three years after diagnosis was 63% (95% CI: 58%, 68%) for the overall cohort. 
- When stratified by gender, males exhibited a higher survival rate at 69% (95% CI: 64%, 75%). 
- Similarly, among different ethnic groups, Indian patients displayed the highest survival rate at 86% (95% CI: 76%, 97%). 
- Additionally, patients diagnosed with ischemic stroke had a survival rate of 63% (95% CI: 59%, 68%), surpassing their counterparts. 

# Prediction of mortality among stroke patients

## The comparison between COX-GCS, COX-NIHSS, and COX-MRS models in predicting stroke mortality

- We perform comparative analysis to evaluate the prognostic value of each stroke severity scales by using confirmatory analysis. 
- Ininitally, we fit an unadjusted cox model for each stroke severity scale, i.e., the GCS, NIHSS, and mRS. 
- Next, we fit an adjusted cox model for sociodemographic characteristics, co morbidities and intervention

```{r, include=FALSE}
library(forcats)
#ss
fct_count(str_mod$ss)
str_mod$ss2 <- fct_collapse(str_mod$ss,
                               NCR=c("PPUKM", "HSP"),
                               ECR=c("HRPZ", "HSNZ", "HUSM"))
str_mod$ss2 <- relevel(str_mod$ss2, ref = "NCR")
fct_count(str_mod$ss2)

```


```{r, include=FALSE}
## Prepare dataset for GCS, NIHSS and mRS

# Prepare GCS dataset

str_gcs <- str_mod %>% gtsummary::select(age, sex, married, ethnicity, ss2, dm, hpt, ckd, af, hf_ihd, lipid, smoke, who, gcs_reduct, iv_thrombolysis, iv_thrombectomy, dur_month, status)

str_nihss <- str_mod %>% gtsummary::select(age, sex, married, ethnicity, ss2, dm, hpt, ckd, af, hf_ihd, lipid, smoke, who, nihss, iv_thrombolysis, iv_thrombectomy, dur_month, status)

str_mrs <- str_mod %>% gtsummary::select(age, sex, married, ethnicity, ss2, dm, hpt, ckd, af, hf_ihd, lipid, smoke, who, mrs, iv_thrombolysis, iv_thrombectomy, dur_month, status)
```

```{r, include=FALSE}

## GCS Cox Model

### Unadjusted model

coxgcs_uv <- coxph(Surv(dur_month, status)~ gcs_reduct, str_gcs)%>% 
  tbl_regression(exponentiate = TRUE, 
                           label = list(
                                 gcs_reduct ~ "Glasgow coma scale (GCS) reduction"
                                 )) 

coxgcs_uv
```

```{r, include=FALSE}
### Adjusted model

coxgcs_mv <- coxph(Surv(dur_month, status)~ gcs_reduct+age+sex+married+ethnicity+ss2+dm+hpt+ckd+af+hf_ihd+lipid+smoke+who+iv_thrombolysis+iv_thrombectomy, str_gcs) %>% 
  tbl_regression(exponentiate = TRUE, 
                             include = "gcs_reduct",
                           label = list(
                                 gcs_reduct ~ "Glasgow coma scale (GCS) reduction")) 

coxgcs_mv

```

```{r, include=FALSE}
gcs_merge <- tbl_merge(list(coxgcs_uv, coxgcs_mv),
                       tab_spanner = c("**Unadjusted model**", "**Adjusted model**"))
gcs_merge
```


```{r, include=FALSE}
## NIHSS Cox Model

### Unadjusted model
coxnihss_uv <- coxph(Surv(dur_month, status)~ nihss, str_nihss) %>% 
  tbl_regression(exponentiate = TRUE, 
                           label = list(
                                 nihss ~ "NIH stroke scale (NIHSS)"
                                 )) 
coxnihss_uv
```


```{r, include=FALSE}
### Adjusted model
coxnihss_mv <- coxph(Surv(dur_month, status)~ nihss+age+sex+married+ethnicity+ss2+dm+hpt+ckd+af+hf_ihd+lipid+smoke+who+iv_thrombolysis+iv_thrombectomy, str_nihss) %>% 
  tbl_regression(exponentiate = TRUE, 
                 include = "nihss",
                           label = list(
                                 nihss ~ "NIH stroke scale (NIHSS)")) 

coxnihss_mv
```

```{r, include=FALSE}
nihss_merge <- tbl_merge(list(coxnihss_uv, coxnihss_mv),
                       tab_spanner = c("**Unadjusted model**", "**Adjusted model**"))
nihss_merge
```


```{r, include=FALSE}
## mRS Cox Model

### Unadjusted model
coxmrs_uv <- coxph(Surv(dur_month, status)~ mrs, str_mrs) %>% 
  tbl_regression(exponentiate = TRUE, 
                           label = list(
                                 mrs ~ "Modified rankin scale (mRS)"
                                 )) 
coxmrs_uv
```


```{r, include=FALSE}
### Adjusted model

coxmrs_mv <- coxph(Surv(dur_month, status)~ mrs+age+sex+married+ethnicity+ss2+dm+hpt+ckd+af+hf_ihd+lipid+smoke+who+iv_thrombolysis+iv_thrombectomy, str_mrs) %>% 
  tbl_regression(exponentiate = TRUE, 
                 include = "mrs",
                           label = list(
                                 mrs ~ "Modified rankin scale (mRS)")) 

coxmrs_mv
```

```{r, include=FALSE}
mrs_merge <- tbl_merge(list(coxmrs_uv, coxmrs_mv),
                       tab_spanner = c("**Unadjusted model**", "**Adjusted model**"))
mrs_merge
```

## Combine all modell together with tbl_stack


```{r, echo=FALSE}
model_stack <- tbl_stack(list(gcs_merge, nihss_merge, mrs_merge))
model_stack %>% 
  bold_labels() %>% 
  italicize_levels()
```

- Regarding the Glasgow Coma Scale (GCS), patients diagnosed with one-unit lower GCS exhibited an 8% higher risk of mortality from stroke (HR=1.08; 95% CI: 1.06, 1.11; p<0.001). 
- An increase in stroke severity based on the NIHSS scale corresponded to a 4.25-fold higher hazard of mortality (95% CI: 2.41, 7.49; p<0.001). 
- In the case of mRS scores upon discharge, a score of five was associated with an 8.28-fold higher risk of stroke-related mortality (95% CI: 2.61, 26.3; p<0.001). 
- Notably, these associations remained significant even after adjusting for sociodemographic factors, comorbidities, and interventions

# Evaluation of the model performace

```{r, include=FALSE}
#NON GT

coxgcs_mv_ngt <- coxph(Surv(dur_month, status)~ gcs_reduct+age+sex+married+ethnicity+ss2+dm+hpt+ckd+af+hf_ihd+lipid+smoke+who+iv_thrombolysis+iv_thrombectomy, str_gcs)

coxnihss_mv_ngt <- coxph(Surv(dur_month, status)~ nihss+age+sex+married+ethnicity+ss2+dm+hpt+ckd+af+hf_ihd+lipid+smoke+who+iv_thrombolysis+iv_thrombectomy, str_nihss)

coxmrs_mv_ngt <- coxph(Surv(dur_month, status)~ mrs+age+sex+married+ethnicity+ss2+dm+hpt+ckd+af+hf_ihd+lipid+smoke+who+iv_thrombolysis+iv_thrombectomy, str_mrs)
```


Now we will evaluate the performance of the model with likelihood based test, C-index and visualization of the fitted model

```{r, include=FALSE}
## Likelihood based test

### Likelihood ratio test
library(lmtest)
lrt_gcs_nihss <- lrtest(coxgcs_mv_ngt, coxnihss_mv_ngt)
lrt_gcs_mrs <- lrtest(coxgcs_mv_ngt, coxmrs_mv_ngt)

lrt_gcs_nihss
lrt_gcs_mrs
```


```{r, include=FALSE}
aic_values <- AIC(coxgcs_mv_ngt, coxnihss_mv_ngt, coxmrs_mv_ngt)
aic_values
```


```{r, include=FALSE}
# C-index

ctest <- concordance(coxgcs_mv_ngt, coxnihss_mv_ngt, coxmrs_mv_ngt)
ctest
```


| Cox Proportional Hazard Model |                                   | Univariable |            |         | Multivariable |            |         | Likelihood ratio test | Akaike Information Criterion |    | Harrell's C-statistic |       |
|-------------------------------|-----------------------------------|-------------|------------|---------|---------------|------------|---------|-----------------------|------------------------------|----|-----------------------|-------|
|                               |                                   | **HR**          | **95% CI**     | **p-value** | **HR**            | **95% CI**     | **p-value** | **p-value**               | **AIC**                          | **DF** | **C-index**               | **SE**    |
| **GCS**                           |                                   | 1.08        | 1.06, 1.11 | <0.001  | 1.25          | 1.20, 1.31 | <0.001  | Reference             | 3285.25                      | 19 | 0.75                  | 0.017 |
| **NIHSS**                         |                                   |             |            |         |               |            |         | 0.002                 | 3276.82                      | 22 | 0.76                  | 0.015 |
|                               | *No stroke symptoms (0)*            | Reference   |            |         | Reference     |            |         |                       |                              |    |                       |       |
|                               | *Minor stroke (1-4)*                | 0.49        | 0.27, 0.88 | 0.017   |      0.50     | 0.27, 0.91 |  0.025  |                       |                              |    |                       |       |
|                               | *Moderate stroke (5-15)*            | 1.32        | 0.77, 2.27 | 0.3     |      1.24     | 0.70, 2.20 |   0.5   |                       |                              |    |                       |       |
|                               | *Moderate to severe stroke (16-20)* | 2.26        | 1.27, 4.03 | 0.006   |      3.04     | 1.63, 5.69 |  <0.001 |                       |                              |    |                       |       |
|                               | *Severe stroke (21-42)*             | 4.25        | 2.41, 7.49 | <0.001  | 3.79          | 2.08, 6.91 | <0.001  |                       |                              |    |                       |       |
| **mRS**                           |                                   |             |            |         |               |            |         | **<0.001**                | **3200.97**                      | 24 | **0.79**                  | 0.015 |
|                               | *mRS score 0*                       | Reference   |            |         | Reference     |            |         |                       |                              |    |                       |       |
|                               | *mRS score 1*                       | 1.01        | 0.29, 3.47 | >0.9    | 0.77          | 0.22, 2.69 | 0.7     |                       |                              |    |                       |       |
|                               | *mRS score 2*                       | 1.71        | 0.51, 5.70 | 0.4     | 1.28          | 0.38, 4.29 | 0.7     |                       |                              |    |                       |       |
|                               | *mRS score 3*                       | 2.85        | 0.88, 9.25 | 0.082   | 1.77          | 0.54, 5.81 | 0.3     |                       |                              |    |                       |       |
|                               | *mRS score 4*                       | 4.20        | 1.32, 13.3 | 0.015   | 2.97          | 0.93, 9.50 | 0.067   |                       |                              |    |                       |       |
|                               | *mRS score 5*                       | 8.28        | 2.61, 26.3 | <0.001  | 5.46          | 1.69, 17.6 | 0.005   |                       |                              |    |                       |       |
|                               | *mRS score 6*                       | 14.2        | 4.41, 45.5 | <0.001  | 41.4          | 12.4, 139  | <0.001  |                       |                              |    |                       |       |

Table 4 The Cox Proportional Hazard regression models with likelihood-based tests, Akaike information criterion, and Harrel's C-statistic.

- Regarding the Glasgow Coma Scale (GCS), patients diagnosed with one-unit lower GCS exhibited an 8% higher risk of mortality from stroke (HR=1.08; 95% CI: 1.06, 1.11; p<0.001). 
- An increase in stroke severity based on the NIHSS scale corresponded to a 4.25-fold higher hazard of mortality (95% CI: 2.41, 7.49; p<0.001). 
- In the case of mRS scores upon discharge, a score of five was associated with an 8.28-fold higher risk of stroke-related mortality (95% CI: 2.61, 26.3; p<0.001). 
- Notably, these associations remained significant even after adjusting for sociodemographic factors, comorbidities, and interventions (see Table 4).


- To compare the predictive performance of stroke mortality using the three scales collected during diagnosis, we conducted a likelihood-based test, computed the C-index, and generated Kaplan-Meier curves. 
- The Cox proportional hazard model based on the mRS scale outperformed the models based on the other scales, with a p-value <0.001, lower AIC (Akaike Information Criterion) at 3200.97, and a higher C-index of 0.79, respectively. 

```{r, echo=FALSE, message=FALSE, warning=FALSE}
## visualize model comparison

# Load the required libraries
library(survival)
library(ggplot2)

# Create Kaplan-Meier survival curves for each model with confidence intervals
surv1 <- survfit(coxgcs_mv_ngt)
surv2 <- survfit(coxnihss_mv_ngt)
surv3 <- survfit(coxmrs_mv_ngt)

# Combine the survival data into a data frame
surv_data <- data.frame(
  time = surv1$time,
  surv1 = surv1$surv,
  surv2 = surv2$surv,
  surv3 = surv3$surv,
  upper1 = surv1$upper,
  lower1 = surv1$lower,
  upper2 = surv2$upper,
  lower2 = surv2$lower,
  upper3 = surv3$upper,
  lower3 = surv3$lower
)

# Create the Kaplan-Meier plot using ggplot2
p <- ggplot(surv_data, aes(x = time)) +
  geom_step(aes(y = surv1, color = "Cox GCS model", fill = "Cox GCS model"), size = 1) +
  geom_step(aes(y = surv2, color = "Cox NIHSS model", fill = "Cox NIHSS model"), size = 1) +
  geom_step(aes(y = surv3, color = "Cox mRS model", fill = "Cox mRS model"), size = 1) +
  geom_ribbon(aes(ymin = lower1, ymax = upper1, fill = "Cox GCS model"), alpha = 0.2) +
  geom_ribbon(aes(ymin = lower2, ymax = upper2, fill = "Cox NIHSS model"), alpha = 0.2) +
  geom_ribbon(aes(ymin = lower3, ymax = upper3, fill = "Cox mRS model"), alpha = 0.2) +
  scale_color_manual(values = c("Cox GCS model" = "blue", "Cox NIHSS model" = "red", "Cox mRS model" = "green")) +
  scale_fill_manual(
    values = c("Cox GCS model" = "blue", "Cox NIHSS model" = "red", "Cox mRS model" = "green"),
    guide = guide_legend(title = "(95% confidence interval)")
  ) +
  labs(
    title = "",
    x = "Time (in months)",
    y = "Survival Probability"
  ) +
  theme_minimal() +
  guides(color = guide_legend(title = "Cox Proportional Hazard Models", override.aes = list(fill = NA)))

p

```


- Additionally, the Kaplan-Meier curve for the mRS model displayed a less steep decline in survival as compared to the other models.

